### 2014-07-01 paper/abstract.tex
paper mods

%!TEX root = series of tubes.tex
\begin{abstract}
3D printers offer extraordinary flexibility for prototyping the shape and mechanical function of objects.  We investigate how 3D models can be modified to facilitate the creation of \emph{interactive} objects that offer dynamic input and output.
We introduce a general technique for supporting the rapid prototyping of interactivity by removing interior material from 3D models to form internal pipes. We describe this new design space of pipes for interaction design, where variables include openings, path constraints, topologies, and inserted media.  We then present \systemnamenospace, a tool for routing such pipes through the interior of 3D models, integrated within a 3D modeling program.  We use two distinct routing algorithms.  The first has users to define pipes' terminals, and uses path routing and physics-based simulation to minimize pipe bending energy, allowing easy insertion of media post-print.  The second allows users to supply a desired internal shape to which we fit a pipe route: for this we developed a novel graph-based routing algorithm.  We present several prototypes created using our tool to show its flexibility and potential.
\end{abstract}

>>>>

%!TEX root = series of tubes.tex
\begin{abstract}
3D printers offer extraordinary flexibility for prototyping the shape and mechanical function of objects.  We investigate how 3D models can be modified to facilitate the creation of \emph{interactive} objects offering dynamic input and output.
We introduce a general technique to support rapidly prototyping interactivity by removing interior material from 3D models to form internal pipes. We describe the design space of pipes for interaction design, where variables include openings, path constraints, topologies, and inserted media.  We then present \systemnamenospace, a tool for routing internal pipes through 3D models, integrated within a 3D modeling program.  We use two distinct routing algorithms.  The first has users define pipes' terminals and uses path routing and physics-based simulation to minimize pipe bending energy, allowing easy insertion of media post-print.  The second lets users supply a desired internal shape to which it fits a pipe route: for this we developed a novel graph-based routing algorithm.  We present prototypes created using our tool showing its flexibility and potential.
\end{abstract}

### 2014-07-01 paper/designspace.tex
making changes thanks to reviews
paper mods

\section{The Design Space of Pipes}
Our review of prior art indicates the potential benifits of using interior pipes for adding interactivity to 3D printed objects, although no general-purpose tools currently exist for this purpose. To guide the design of such a tool, we now provide a design space of the types of pipes that a maker may want to utilize. We have identified four relevant dimensions of pipe design: \emph{openings}, \emph{path constraints}, \emph{topologies}, and inserted \emph{media}.  We describe the space (Figure \ref{fig:pipespace}).

@@

%\begin{figure}
%\centering
%    \includegraphics[width=1.0\columnwidth]{figures/glamour/return.jpg}
%\caption{A path-constrained open pipe with EL wire as a ``neon'' design.}
%\label{fig:justneon}
%\end{figure}

\subsection{Path Constraints}
Depending on the purpose of the pipe, its geometry may be constrained to specific locations within the 3D model. An \emph{endpoint-constrained} pipe requires specific start and end points, but the particular route of the internal path is irrelevant. This is the case for circuit schematics, where the electronic components have fixed locations, but many routing solutions are valid. In contrast, for a \emph{path-constrained} pipe the shape of the path is important. For example, threading electroluminescent (EL) wire through a clear pipe enables makers to emulate neon art (Figure \ref{fig:teaser}, right). A \emph{fully constrained} pipe requires specific terminals as well as a specific path. If a maker wishes the EL wire to exit the rear of the sign to hide the electronics, she can use a fully-constrained path (Figure \ref{fig:teaser}, right). Finally, an \emph{unconstrained} pipe does not need a specific path or terminal points. A cavity to modify an object's weight is an example of an unconstrained pipe (Figure \ref{fig:openings}c).

@@

\subsection{Media in Pipes}
After printing, different media can be inserted into pipes to change affordances and capabilities. We consider \emph{gas}, \emph{liquids}, \emph{solids}, \emph{particulates}, and \emph{threadables}. 
%
Compressible fluids (\emph{``gas''}) and incompressible fluids (\emph{``liquids''})  inside semi-closed pipes can create haptic feedback; pressure can also be sensed to act as an input~\cite{Slyper-shape}. Gases can be used as carriers for scents or fog. Conductive liquids like copper paint, when dried, turn pipes into conductive paths and can also fix electronic components in place.
%
Filling pipes with \emph{solids} at print time enables designers to exploit the difference in material properties between pipe and contents. For example, in Printed Optics~\cite{Willis-printedoptics}, different refractive indices between solid cladding material and transparent material inside pipes allow light transport along pipes for input and output.
%
\emph{Particulates}, either printed in-place or inserted, can be of varying densities and can contain particles of various and variable size.  A single large particle can be used for display.  Small particles in variable amounts of fluid can provide variable stiffness haptics via jamming \cite{Follmer-jamming}.
%
\emph{Threadable} media, such as regular wire, electroluminescent wire, or commercial fiberoptic cables, can be threaded through pipes post-printing. Threading such media enables addition of conduction, illumination, or optical sensing using materials that can easily be procured, but cannot themselves be printed, e.g., on consumer printers with a single material. 
% This empowers inexpensive printers: for example, a Printed Optics-style interface can be created on a consumer-grade 3D printer with pipes and fiber optic cable.

>>>>

\section{The Design Space of Pipes}
Our review of prior art indicates the potential benefits of using interior pipes for adding interactivity to 3D printed objects, although no general-purpose tools currently exist for this purpose. To guide the design of such a tool, we now provide a design space of the types of pipes that a maker may want to utilize. We have identified four relevant dimensions of pipe design: \emph{openings}, \emph{path constraints}, \emph{topologies}, and inserted \emph{media}.  We describe the space in Figure \ref{fig:pipespace}.

@@

\subsection{Path Constraints}
Depending on the purpose of the pipe, its geometry may be constrained to specific locations within the 3D model. An \emph{endpoint-constrained} pipe requires specific start and end points, but the particular route of the internal path is irrelevant. This is the case for circuit schematics, where the electronic components have fixed locations, but many routing solutions are valid. In contrast, for a \emph{path-constrained} pipe the shape of the path is important. For example, threading electroluminescent (EL) wire through a clear pipe enables makers to emulate neon art (Figure \ref{fig:teaser}, right). A \emph{fully constrained} pipe requires specific terminals as well as a specific path. If a maker wishes the EL wire to exit the rear of the sign to hide the electronics, she can use a fully-constrained path (Figure \ref{fig:teaser}, right). Finally, an \emph{unconstrained} pipe does not need a specific path or terminal points. A cavity to modify an object's overall weight is an example of an unconstrained pipe (Figure \ref{fig:openings}c).

@@

\subsection{Media in Pipes}
After printing, different media can be inserted into pipes to change affordances and capabilities. We consider \emph{gas}, \emph{liquids}, \emph{solids}, \emph{particulates}, and \emph{threadables}. 
%
Compressible fluids (\emph{``gas''}) and incompressible fluids (\emph{``liquids''})  inside semi-closed pipes can create haptic feedback; pressure can also be sensed to act as an input~\cite{Slyper-shape}. Gases can be used as carriers for scents or fog. Conductive liquids like copper paint, when dried, turn pipes into conductive paths and can also fix electronic components in place.
%
Filling pipes with \emph{solids} at print time enables designers to exploit the difference in material properties between pipe and contents. For example, in Printed Optics~\cite{Willis-printedoptics}, different refractive indices between solid cladding material and transparent material inside pipes allow light transport along pipes for input and output.
%
\emph{Particulates}, either printed in-place or inserted, can be of varying densities and can contain particles of various and variable size.  A single large particle can be used for display.  Small particles agitated by air can generate sound.
%
\emph{Threadable} media, such as regular wire, electroluminescent wire, or commercial fiberoptic cables, can be threaded through pipes post-printing. Threading such media enables addition of conduction, illumination, or optical sensing using materials that can easily be procured, but cannot themselves be printed, e.g., on consumer printers with a single material. 
% This empowers inexpensive printers: for example, a Printed Optics-style interface can be created on a consumer-grade 3D printer with pipes and fiber optic cable.

### 2014-07-01 paper/discussion.tex
paper mods

We are also eager to test our tool with users and get feedback from makers who use it in the wild. Based on our own experiences, we believe our system could be readily utilized by makers and researchers, and serve as a useful tool for rapdily proptotyping interactive 3D objects.

>>>>

We are also eager to test our tool with users and get feedback from makers who use it in the wild. Based on our own experiences, we believe our system could be readily utilized by makers and researchers, and serve as a useful tool for rapidly prototyping interactive 3D objects.  \changes{We plan to release PipeDream as a part of Meshmixer later this year.}

### 2014-07-01 paper/introduction.tex
paper mods

Recently, human-computer interaction researchers have begun to explore how to prototype interactive objects using additive 3D printing processes.
Existing approaches, including Printed Optics \cite{Willis-printedoptics} and Sauron \cite{Savage-sauron}, exclusively focus on optical sensing and visual displays --- we seek to grow the vocabulary of supported tangible input and output modalities.

This paper introduces a novel tool that provides path planning and other high-level functionality to support the design of interior pipes and cavities within 3D models.  These pipes can be filled, post-print, with a variety of media to enable input, display, and tactile feedback.  This {\em subtractive} approach is complementary to {\em additive} approaches that inject custom print materials, such as conductors, during the actual printing process \cite{Sells-reprap}. While our approach requires some manual assembly after a print completes, it gives makers the opportunity to cheaply and rapidly prototype a range of interactive objects on 3D printers, including popular single-material fused-deposition modeling machines (colored objects in Figure \ref{fig:teaser}).

We describe a new design space of pipes for the purpose of interaction design, where variables include openings, path constraints, topologies, and inserted media. By exploring this design space, new opportunities for adding interactivity to 3D printed objects can be realized. For example, injecting conductive paint can turn pipes into conductors, which can in turn integrate electronic components into prints; electroluminescent (EL) wire can be threaded through pipes to create computer-controlled visual output; or air can be pumped through pipes to produce haptic effects.  Existing sensing and actuation approaches, such as FlyEye \cite{Wimmer-flyeye}, Swept Frequency Capacitive Sensing \cite{Sato-touche}, and Jamming User Interfaces \cite{Follmer-jamming}, can be expanded to new contexts using pipes: makers utilizing such I/O strategies can locate input and output at arbitrary points on a printed 3D object's surface.  We leverage these existing techniques for sensing and actuation while our work's novelty is in \emph{internal modeling} for 3D printing and the \emph{redirection} of I/O to or through arbitrary locations on a device.

>>>>

Recently, human-computer interaction researchers have begun to explore how to prototype interactive objects using additive 3D printing processes.
Existing approaches, including Printed Optics \cite{Willis-printedoptics} and Sauron \cite{Savage-sauron}, exclusively focus on optical sensing and visual displays --- we seek to grow the vocabulary of supported tangible input and output modalities, and we propose using pipes and cavities, allowing designers to integrate secondary materials post-print, for this purpose.

\changes{This paper introduces a novel tool that provides path planning and other high-level functionality to support the design of interior pipes and cavities within 3D models.  These pipes can be filled, post-print, with a variety of media to enable input, display, and tactile feedback.}  This {\em subtractive} approach is complementary to {\em additive} approaches that inject custom print materials, such as conductors, during the actual printing process \cite{Sells-reprap}. While our approach requires some manual assembly after a print completes, it gives makers the opportunity to cheaply and rapidly prototype a range of interactive objects on 3D printers, including popular single-material fused-deposition modeling machines (colored objects in Figure \ref{fig:teaser}).  \changes{Additionally, once printers able to deposit conductive materials become commercially available, PipeDream will still support the internal 3D modeling tasks required for directly printing different materials.}

We describe a new design space of pipes for the purpose of interaction design, where variables include openings, path constraints, topologies, and inserted media. By exploring this design space, new opportunities for adding interactivity to 3D printed objects can be realized. For example, injecting conductive paint can turn pipes into conductors, which can in turn integrate electronic components into prints; electroluminescent (EL) wire can be threaded through pipes to create computer-controlled visual output; or air can be pumped through pipes to produce haptic effects.  Existing sensing and actuation approaches, such as FlyEye \cite{Wimmer-flyeye}, swept-frequency capacitive sensing \cite{Sato-touche}, and Jamming User Interfaces \cite{Follmer-jamming}, can be expanded to new contexts using pipes: makers utilizing such I/O strategies can locate input and output at arbitrary points on a printed 3D object's surface.  We leverage these existing techniques for sensing and actuation while our work's novelty is in \emph{internal modeling} for 3D printing and the \emph{redirection} of I/O to or through arbitrary locations on a device.

### 2014-07-01 paper/relatedwork.tex
paper mods

Sauron~\cite{Savage-sauron} tracks interactions with printed physical mechanisms by placing a camera inside hollow objects, where all components must be visible to the camera.  Our technique instead can work with arbitrary, solid models and only carves out pipes for interaction where needed. Midas~\cite{Savage-midas} generates fabrication files for custom capacitive touch sensors.  Their technique is limited to planar and unfoldable shapes as it routes connections on the surface of objects. \systemname can also be used for touch sensing, but routes conductors through the inside of objects.

@@

\subsubsection{Fabrication with Printed Electronics}
Researchers are investigating how to print conductors and place electronic components as part of additive manufacturing.  Sells, et al., \cite{Sells-reprap} use a heated syringe to deposit metal in printed 2D circuitboard channels on objects' surfaces.  Other experiments done by the RepRap project\footnote{\url{http://reprap.org}} use plastic filament blended with conductive materials, however the resistance is currently too high to be useful in circuitry.  Sarik, et al., \cite{Sarik-tracebrush} use aerosol jetting of conductive material onto printed objects.  Park, et al., \cite{Park-microchannels} and Majidi, et al., \cite{Majidi-curvature} inject liquid metals.   Our approach does not deposit such materials automatically, but creates structures inside a model that make manual post-print injection of conductive ink possible.

\subsubsection{Interaction for Fabrication vs. Fabrication for Interaction}
HCI researchers are also developing novel interaction techniques for providing input to digital fabrication machines~\cite{Mueller-constructable,mixfab,Willis-interactive}, and new hybrid hand tools that combine manual processes with digital fabrication~\cite{rivers2012position,zoran2013freed}. These projects are complementary as they target different aspects of the design process, though it is conceivable that our pipe design tools be added to AR modeling interfaces like MixFab~\cite{mixfab}.

>>>>

Sauron~\cite{Savage-sauron} tracks interactions with printed physical mechanisms by placing a camera inside hollow objects, where all components must be visible to it.  Our technique instead can work with arbitrary, solid models and only carves pipes for interaction where needed. Midas~\cite{Savage-midas} generates fabrication files for custom capacitive touch sensors.  Their technique is limited to planar and unfoldable shapes, as it routes connections on the surface of objects. \systemname can also be used for touch sensing, but routes conductors \emph{through} objects.

@@

\subsubsection{Fabrication with Printed Electronics}
Researchers are investigating how to print conductors and place electronic components as part of additive manufacturing.  Sells, et al., \cite{Sells-reprap} use a heated syringe to deposit metal in printed 2D circuitboard channels on objects' surfaces.  Other experiments done by the RepRap project\footnote{\url{http://reprap.org}} use plastic filament blended with conductive materials, however the resistance is currently too high to be useful in circuitry.  Sarik, et al., \cite{Sarik-tracebrush} use aerosol jetting of conductive material onto printed objects.  Park, et al., \cite{Park-microchannels} and Majidi, et al., \cite{Majidi-curvature} inject liquid metals, \changes{and Hudson \cite{Hudson-bears} provides techniques for embedding conductive thread and other electronics within felted objects. }  Our approach does not deposit these materials automatically, but creates structures inside a model that make manual post-print injection of conductive ink possible.

\subsubsection{Interaction for Fabrication vs. Fabrication for Interaction}
HCI researchers are also developing novel interaction techniques for providing input to digital fabrication machines~\cite{Mueller-constructable,mixfab,Willis-interactive}, \changes{new applications to extend machine capabilities \cite{Mueller-laserorigami,Hudson-bears}}, and new hybrid hand tools that combine manual processes with digital fabrication~\cite{rivers2012position,zoran2013freed}. These projects are complementary as they target different aspects of the design process, though it is conceivable that our pipe design tools be added to AR modeling interfaces like MixFab~\cite{mixfab}.

### 2014-07-01 paper/tubes.tex
paper mods

% Designing objects which, for example, are touch sensitive in particular areas or contain electronic components, requires precise location and sizing of pipe endpoints.  The interior of these pipes should have as large a bending radius as possible so that post-print insertion of solid or viscous media is easier.
To create endpoint-constrained pipes, in which the location of exterior connection points matters, a brush tool is used to specify a pair of points on the mesh's surface. The system automatically routes the pipe, and displays a preview.  Using sliders, the user can independently adjust the starting and ending radii of the pipe.  This is useful for, e.g, our breathing bunny (Figure \ref{fig:examples}b), whose pipes must fit a piece of hardware on one end, but on the other end should have as large an interactive surface as possible.

@@

\subsubsection{Routing Algorithm and Physical Simulation}
Given two points, we first find a path between them using the A* search algorithm on a voxel representation of the object~\cite{Hart-Astar}. A* finds a least-cost path through a graph of nodes given a cost function. Our graph comprises the grid of voxels as nodes, with edges between adjacent nodes; our path cost is based on the Euclidean distance to the selected end point. In this formulation, the routed path is naturally constrained to stay within the object boundaries.
%; for each point it tags the distance from the start (e.g., startPoint = 0, startNeighbors=1, ...).  Once the end point is reached, we backtrack through tagged points, always moving to a lower-valued point. \bjoern{find a clearer way to express this?}

@@

We solve this problem via physical simulation. We create a straight virtual 
wire (a 3D polyline or \emph{discrete rod}), and then set its initial position 
as our A* path.
Running the physical simulation allows the wire to attempt to return to
a straight configuration, while satisfying the user-selected endpoint position constraints.  In addition to user constraints, we require that the rod be perpendicular to the surface at its ends to prevent the terminals' shape from deforming.

@@

\begin{figure}[h!]
\centering
    \includegraphics[width=3.4in]{figures/routing-tests.png}
\caption{A set of example routings produced by our tool.  Note that because the initial state of the discrete rod is based on an A* routing, it cannot jump around holes in a genus $>0$ mesh (a) even if that would reduce overall bending.  Pipes successfully route through long, curving paths in knots (b), and multiple pipes placed in a single mesh in a tangled configuration successfully avoid each other (c).}
\label{fig:complex}
\end{figure}

@@

\begin{figure}
\centering
    \includegraphics[width=3.4in]{figures/radius.png}
\caption{Users can interactively change the radius of their pipes using a slider (b).  PipeDream's live preview shows the real size and location of the cut before it is committed to the mesh.}
\label{fig:radius}
\end{figure}

@@

In some cases, such as for the neon sign in Figure \ref{fig:tool-process-interior}, a single path must pass through a set of disconnected path segments (i.e., all letters in the word ``UIST'' ).  \systemname leverages graph theory algorithms to generate this single path, as described below.  Without endpoint constraints, the path's exit(s) from the model are defined by the SVG: the maker can see where the path exits the model and translate the cut according to her desired outcome.

@@

\section{Fabrication Techniques}
Several pragmatic concerns arise when fabricating models with pipes and cavities. Pipes frequently lead to overhangs (where an underlying supporting layer has a smaller footprint than the subsequent layer).
To allow overhangs, 3D printers may lay down support material that must be later removed from the model.  The removal process can be time-intensive and challenging.  The physical fabrication process used by different machines gives rise to differing strategies for avoiding support material or easing its removal.  Techniques differ for fused deposition modeling (FDM) machines, such as the Makerbot\footnote{\url{http://makerbot.com}}, polyjet machines such as Stratasys Connex\footnote{\url{http://stratasys.com}},  powder-based machines (e.g., ZCorp\footnote{\url{http://3dsystems.com}}), and those that use stereolithography (e.g., Form 1\footnote{\url{http://formlabs.com}}).

@@

\subsection{Model Assembly}
For especially complex geometries, models can be ``cut up'' into multiple pieces that can be assembled, avoiding support deposition or easing material removal; this can also ease insertion of the desired medium. This is somewhat undesirable as it means more time spent in assembly post-print, and additionally a cut pipe may no longer be fluid-tight.  Our plane cut option goes towards this, however a more advanced tool for this purpose is outside the scope of this work.

>>>>

% Designing objects which, for example, are touch sensitive in particular areas or contain electronic components, requires precise location and sizing of pipe endpoints.  The interior of these pipes should have as large a bending radius as possible so that post-print insertion of solid or viscous media is easier.
To create endpoint-constrained pipes, in which the location of exterior connection points matters, a brush tool is used to specify a pair of points on the mesh's surface (Figure \ref{fig:tool-process-exterior}a). The system automatically routes the pipe, and displays a preview.  Using sliders, the user can independently adjust the starting and ending radii of the pipe.  This is useful for, e.g, our breathing bunny (Figure \ref{fig:examples}b), whose pipes must fit a piece of hardware on one end, but on the other end should have as large an interactive surface as possible.

@@

\subsubsection{Routing Algorithm and Physical Simulation}
Given two points, we first find a path between them using the A* search algorithm on a voxel representation of the object~\cite{Hart-Astar}. A* finds a least-cost path through a graph of nodes given a cost function. Our graph comprises the grid of voxels as nodes, with edges between adjacent nodes; our path cost is based on the Euclidean distance to the selected end point. In this formulation, the routed path is naturally constrained to stay within the object boundaries (Figure \ref{fig:tool-process-exterior}b).
%; for each point it tags the distance from the start (e.g., startPoint = 0, startNeighbors=1, ...).  Once the end point is reached, we backtrack through tagged points, always moving to a lower-valued point. \bjoern{find a clearer way to express this?}

@@

We solve this problem via physical simulation. We create a straight virtual 
wire (a 3D polyline or \emph{discrete rod}), and then set its initial position 
as our A* path.
Running the physical simulation allows the wire to attempt to return to
a straight configuration, while satisfying the user-selected endpoint position constraints.  In addition to user constraints, we require that the rod be perpendicular to the surface at its ends to prevent the terminals' shape from deforming (Figure \ref{fig:tool-process-exterior}c).

@@

\begin{figure}[h!]
\centering
    \includegraphics[width=3.4in]{figures/routing-tests.png}
\caption{\changes{A set of example routings produced by our tool.  Note that because the initial state of the discrete rod is based on an A* routing, it cannot ``jump around'' holes in a genus $>0$ mesh (a) to route on the other side, even if this would reduce overall bending.  Pipes successfully route through long, curving paths in knots (b), and multiple pipes placed in a single mesh in a tangled configuration successfully avoid each other (c).}}
\label{fig:complex}
\end{figure}

@@

\begin{figure}
\centering
    \includegraphics[width=3.4in]{figures/radius.png}
\caption{Users can interactively change the radius of their pipes using a slider (b).  PipeDream's live preview shows the real size and location of the cut (a, c) before it is committed to the mesh.}
\label{fig:radius}
\end{figure}

@@

In some cases, such as for the neon route in Figure \ref{fig:tool-process-interior} and resultant sign in Figure \ref{fig:UIST}, a single path must pass through a set of disconnected path segments (i.e., all letters in the word ``UIST'' ).  \systemname leverages graph theory algorithms to generate this single path, as described below.  Without endpoint constraints, the path's exit(s) from the model are defined by the SVG: the maker can see where the path exits the model and translate the cut according to her desired outcome.

@@

\changes{
\section{Fabrication Techniques}
Several pragmatic concerns arise when fabricating models with pipes and cavities. Pipes frequently lead to overhangs (where an underlying supporting layer has a smaller footprint than the subsequent layer).
To allow overhangs, 3D printers may lay down support material that must be later removed from the model.  The removal process can be time-intensive and challenging.  The physical fabrication process used by different machines gives rise to differing strategies for avoiding support material or easing its removal.  Techniques differ for fused deposition modeling (FDM) machines, such as the Makerbot\footnote{\url{http://makerbot.com}}, polyjet machines such as Stratasys Connex\footnote{\url{http://stratasys.com}},  powder-based machines (e.g., ZCorp\footnote{\url{http://3dsystems.com}}), and those that use stereolithography (e.g., Form 1\footnote{\url{http://formlabs.com}}).

@@

\subsection{Model Assembly}
For especially complex geometries, models can be ``cut up'' into multiple pieces that can be assembled, avoiding support deposition or easing material removal; this can also ease insertion of the desired medium. This is somewhat undesirable as it means more time spent in assembly post-print, and additionally a cut pipe may no longer be fluid-tight.  Our plane cut option goes towards this, however a more advanced tool for this purpose is outside the scope of this work.
}
